---
title: "Bayesian ROS model example"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(here)

library(cmdstanr)
library(posterior)

library(tidybayes)
library(ggplot2)
theme_set( theme_bw() )

MODEL_DIR <- here("stan_models")
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)

```


## Helper code

A utility function for sampling from cmdstanr models. This was provided in Richard McElreath's workshop materials (`script.R`) but has been slightly tweaked here to store the Stan model in specified directory and with a file name based on the name of the variable passed to the `model_code` argument. 

```{r}

get_samples <- function(model_code, data=list(), 
                        model_dir = MODEL_DIR, 
                        model_basename = NULL,
                        seed=123, chains=4) {
  
  if (is.null(model_basename)) {
    model_basename <- deparse(substitute(model_code))
    
    # remove '_code' from the base name if present
    model_basename <- sub(model_basename, pattern = "_code", replacement = "")
  }
  
  f <- write_stan_file(model_code, 
                       dir = model_dir, 
                       basename = model_basename)
  
  model_x <- cmdstan_model(f)
  samples_1 <- model_x$sample(
    data = data,
    seed = seed,
    chains = chains,
    parallel_chains = chains,
    refresh = 0
  )
  pr <- as_draws_rvars( samples_1$draws() )
  p <- list()
  for ( i in 1:length(pr) )
    p[[ names(pr)[i] ]] <- draws_of( pr[[i]] )
  return(p)
}

```

### the data

```{r}

#load 10 spread lines
dat.ros <- readRDS(here("example\\dat.ros.rds"))

#load hourly weather for the 10 spread lines. "lineid" can be used to link weather to spread line
dat.weather <- readRDS(here("example\\dat.weather.rds"))

#summarise weather prior to model as first test.
dat.weather.mean <- dat.weather %>% 
  group_by(lineid) %>% 
  summarise(wind_mean=mean(wind),
            rh_mean=mean(rh))


glimpse(dat.ros)
glimpse(dat.weather.mean)
glimpse(dat.weather)

```


### the model
This model is fitted based of mean sampled weather.
Assuming ros is from normal distribution for simplicity for the moment - the actual results don't matter yet, just learning the stan model structure.
```{r}

model_code <- "
data {
  int<lower=0> N;   // number observations
  vector[N] wnd;           // first predictor - wind
  vector[N] rh;           // second predictor - relative humidity
  vector[N] ros;      // outcome - rate of spread
}
parameters {
  real alpha;           // intercept - No specific priors defined but I could define weakly informative priors in model block
  real beta_wnd;       // coefficients for predictors
  real beta_rh;       // coefficients for predictors
  real<lower=0> sigma;  // error scale
}
model {
    ros ~ normal(alpha + beta_wnd * wnd + beta_rh * rh, sigma);  // likelihood. Assuming normal distrubtion for now, but this will need to change
}"

# compile and sample
post <- get_samples(
  model_code,
  data = list(N=nrow(dat.ros), 
              ros=dat.ros$ros,
              wnd=dat.weather.mean$wind_mean,
              rh=dat.weather.mean$rh_mean)
)
```
plot the coefficients
these won't be very good - N ROS = 10
```{r}
dat_gg <- as.data.frame(post) %>% 
  select(-lp__) 


dat_gg %>% 
  tidyr::pivot_longer(everything()) %>% 
  ggplot(aes(x = value, colour = name)) +
  geom_density(linewidth = 1)+
  #lims(x=c(0,50))+
  facet_wrap(~name,scales = "free")

```

