---
title: "bayesian_ros_model_2"
output: html_document
date: "2025-07-24"
---

# Sigmoid ROS Model in R with cmdstanr

## Load required packages

```{r}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(stringr)
library(here)

library(cmdstanr)
library(posterior)

library(ggplot2)
theme_set( theme_bw() )

MODEL_DIR <- here("stan_models")
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)

```
## Load data
## ROS and weather

Load the ROS and weather data. Select a couple of weather variables to use as predictors for an initial model.

```{r}
dat.ros <- readRDS(here("data/lines.rds")) 
dat.weather <- readRDS(here("data/weatherhourly.rds"))

#select relevant ROS variable - ros and lineid, which matches lineid in weather data
dat.ros <- dat.ros %>% 
  select(lineid,ros=ros_kmh,
         start_time_utc,
         end_time_utc) 

#select weather variables
#mean refers to mean of all barra cells that intersect the spread line.
dat.weather <- dat.weather %>% 
  select(lineid,
         wind=sfcWind_mean,
         rh=hurs_mean,
         time_utc_hourly
  ) %>% 
  mutate(wind=wind*3.6)#kmh


#weather is sampled hourly for the spread time of each line
#add variable to identify if BARRA sample is before, during or after spread time
dat.weather <- dat.weather %>% 
  left_join(dat.ros %>% select(lineid,start_time_utc,end_time_utc)) %>% 
  
  #round start and end times, to match barra times
  mutate(start_time_utc_round=lubridate::round_date(start_time_utc,unit = "hours"),
         end_time_utc_round=lubridate::round_date(end_time_utc,unit = "hours")) %>% 
  
  #remove any sampled barra weather from after end time
  filter(time_utc_hourly <= end_time_utc_round) %>% 
  
  #calculate time difference between barra sample time and start time
  #this may be useful to know, for example, wind speed 1 hour vs 3 hours before spread
  mutate(diff_start_hour=as.numeric(difftime(time_utc_hourly,start_time_utc_round,units = "hours"))) %>% 
  
  select(lineid,wind,rh,diff_start_hour)


#plot some of the weather and ros

dat.ros %>% 
  ggplot() +
  geom_histogram(aes(x=ros))

sample.ids <- sample(1:nrow(dat.ros), size = 9, replace = TRUE)
dat.weather %>% 
  filter(lineid %in% sample.ids) %>% 
  ggplot()+
  geom_density(aes(x=wind))+
  facet_wrap(~lineid)



```

## organise data
prepare data for stan model.
```{r}

# Weather data for fire spread lines
dat <- dat.weather %>%
  filter(diff_start_hour > -2)

# Assemble Stan input list
model_data <- list(
  W = nrow(dat),                     # number of weather records
  L = max(dat$lineid),              # number of lines
  wline_id = dat$lineid,            # which line each weather obs belongs to
  wind = dat$wind,                  # hourly wind values
  rh = dat$rh / 100                 # hourly RH - convert RH to proportion
)

# Add rate of spread values for lines
model_data <- c(model_data, 
                 list(ros_line_id = dat.ros$lineid, ros = dat.ros$ros))

# Check that the weather line IDs and the ROS line IDs line up properly
MaxLineID <- max(model_data$wline_id)

# Check there are no gaps in the line IDs
stopifnot( all(model_data$wline_id %in% seq_len(MaxLineID)) )
  
# Check that all weather line IDs are in the ROS line IDs and vice versa
stopifnot( length(symdiff(model_data$wline_id, model_data$ros_line_id)) == 0)

```

## Helper code

A utility function for sampling from cmdstanr models. This was provided in Richard McElreath's workshop materials (`script.R`) but has been slightly tweaked here to store the Stan model in specified directory and with a file name based on the name of the variable passed to the `model_code` argument. 
Modified to just return samples_1

```{r}

get_samples <- function(model_code, data=list(), 
                        model_dir = MODEL_DIR, 
                        model_basename = NULL,
                        seed=123, chains=4,refresh=0) {
  
    if (is.null(model_basename)) {
      model_basename <- deparse(substitute(model_code))
      
      # remove '_code' from the base name if present
      model_basename <- sub(model_basename, pattern = "_code", replacement = "")
    }
  
    f <- write_stan_file(model_code, 
                         dir = model_dir, 
                         basename = model_basename)
    
    model_x <- cmdstan_model(f)
    samples_1 <- model_x$sample(
        data = data,
        seed = seed,
        chains = chains,
        parallel_chains = chains,
        refresh = 0
    )
    
    return(samples_1)
    # pr <- as_draws_rvars( samples_1$draws() )
    # p <- list()
    # for ( i in 1:length(pr) )
    #     p[[ names(pr)[i] ]] <- draws_of( pr[[i]] )
    # return(p)
}

```


## Model code
ROS model
Sigmoid shaped linear predictor
estimated mean weather for each line from hourly observations.

```{r}
model_code_sigmoid <- "
data {
  int<lower=1> W;
  int<lower=1> L;

  array[W] int<lower=1> wline_id;
  vector<lower=0>[W] wind;
  vector<lower=0>[W] rh;
  vector<lower=0>[L] ros;
}
parameters {
  vector[L] avg_wind;
  real<lower=0> sigma_wind;

  vector<lower=0.01, upper=0.99>[L] avg_rh;
  real<lower=0, upper=50> phi_rh;

  real alpha;
  real beta_wind;
  real beta_rh;

  real<lower=0> shape_ros;
}
model {
  real max_ros = 15;
  vector[L] mu;
  vector[L] rate;

  for (i in 1:W) {
    wind[i] ~ normal(avg_wind[wline_id[i]], sigma_wind);
    rh[i] ~ beta_proportion(avg_rh[wline_id[i]], phi_rh + 1);
  }

  for (i in 1:L) {
    real linpred = alpha + beta_wind * avg_wind[i] + beta_rh * avg_rh[i];
    mu[i] = max_ros / (1 + exp(linpred));
    rate[i] = shape_ros / mu[i];
  }

  ros ~ gamma(shape_ros, rate);
}
"
```

## Get samples


```{r}
#fit. The helper function now returns samples (CmdStan object)
post_sigmoid <- get_samples(
  model_code_sigmoid,
  data = model_data ,
  refresh = 100
)

# Extract tidy posterior draws
post_sigmoid_draws <- as_draws_df(post_sigmoid$draws()) %>% 
  spread_draws(alpha,beta_wind,beta_rh,shape_ros,avg_wind[lineid],avg_rh[lineid],
               
               ndraws = 500)

# Add observed predictors for plotting
dat.obs.sum <- dat.weather %>% 
  group_by(lineid) %>% 
  summarise(wind_obs_mean=mean(wind),
            rh_obs_mean=mean(rh))

post_sigmoid_draws <- post_sigmoid_draws %>% 
  left_join(dat.obs.sum)

```


## predict
Generate predictions for new data

```{r}
#prediction grid
dat_pred <- expand.grid(
  wind = seq(0, 80, 20),
  rh = c(0.2, .8)
)

#posterior predictions
ros_pred <- post_sigmoid_draws %>%
  tidyr::crossing(dat_pred) %>%
  mutate(
    linpred = alpha + beta_wind * wind + beta_rh * rh,
    mu = 10 / (1 + exp(linpred)),
    ros = rgamma(n(), shape = shape_ros, rate = shape_ros / mu)
  )
```


Plot sample of predicted mean values 
```{r}
ros_pred %>% 
  filter(.draw %in% sample(unique(.draw),size=50)) %>% 
  mutate(rh=factor(rh*100)) %>% 
  ggplot(data=.)+
  geom_line(aes(x = wind,y = mu, color = rh, group = paste(rh,.draw)), alpha = 0.9)
```

summarise predicted ROS 

```{r}
ros_pred_sum <- ros_pred %>%
  group_by(wind,rh) %>%
  summarise(
    prc_ros_2=sum(ros > 2)/n() *100,
    ros_median = median(ros),
    ros_lower_90 = quantile(ros, 0.05),
    ros_upper_90 = quantile(ros, 0.95),
    ros_lower_50 = quantile(ros, 0.25),
    ros_upper_50 = quantile(ros, 0.75),
    .groups = "drop"
  ) %>% 
  mutate(rh=factor(rh*100))
```


plot median prediction with intervals
```{r}
ggplot(ros_pred_sum ) +
    geom_ribbon(aes(x = wind, ymin = ros_lower_50, ymax = ros_upper_50,
                    fill = rh), 
               , alpha = 0.3) +
  geom_line(aes(x = wind, y = ros_median, color = rh), linewidth = 1.2) +
  labs(x = "Wind (km/h)", y = "Rate of Spread (km/h)",
       title = "Predicted Rate of Spread vs Wind Speed")




```


boxplots of predictions with interval
```{r}
ros_pred_sum %>% 
  ggplot(aes(x = wind)) +

  geom_linerange(aes(ymin = ros_lower_90, ymax = ros_upper_90, colour = rh),
                 position = position_dodge(width = 5),
                 linewidth = 1) +
  
  geom_linerange(aes(ymin = ros_lower_50, ymax = ros_upper_50, colour = rh),
                 position = position_dodge(width = 5),
                 linewidth = 2.5, size = 1.5) +
  
  labs(colour = "Relative humidity",
       x = "Wind speed (km/h)",
       y = "Predicted ROS")  
```


tile plot of chance prediction ROS > 2

```{r}

ggplot(ros_pred_sum, aes(x = wind, y = rh, fill = prc_ros_2)) +
  geom_tile() +
   geom_text(aes(label = round(prc_ros_2,0)), alpha = 0.7) +
  scale_fill_gradient(low="white",high="red",name = "% > 2kmh") +
  labs(title = "% chance ROS > 2kmh") +
  theme_minimal()
```

## Shiny app 
Shiny app to produce predictions by selecting rh and wind speed
```{r}

library(shiny)
library(ggplot2)
library(dplyr)

ui <- fluidPage(
  titlePanel("Predicted Rate of Spread"),
  
  sidebarLayout(
    sidebarPanel(
      sliderInput("rh", "Relative Humidity (%)", min = 0, max = 100, value = 10),
      sliderInput("wind", "Wind Speed (km/h)", min = 0, max = 100, value = 20)
    ),
    
    mainPanel(
      plotOutput("rosPlot")
    )
  )
)

server <- function(input, output) {
  output$rosPlot <- renderPlot({
    # Create the prediction input
    dat_pred <- expand.grid(
      wind = input$wind,
      rh = input$rh / 100  # convert to proportion
    )
    
    # Simulate ROS using the posterior
ros_pred <- post_sigmoid_draws %>%
  filter(.draw %in% sample(unique(.draw),size=50)) %>% #sample of draws to speed up processing time.
  tidyr::crossing(dat_pred) %>%
  mutate(
    linpred = alpha + beta_wind * wind + beta_rh * rh,
    mu = 10 / (1 + exp(linpred)),
    ros = rgamma(n(), shape = shape_ros, rate = shape_ros / mu)
  )
    
ros_pred_sum <- ros_pred %>% 
  group_by(wind,rh) %>% 
  summarise(ros_2kmh=round(sum(ros>2)/n()*100,0))

ros_2km <- ros_pred_sum$ros_2kmh

glimpse(ros_pred_sum)

    # Plot the density
    ggplot(ros_pred, aes(x = ros)) +
      geom_density(fill = "firebrick", alpha = 0.7) +
      geom_vline(xintercept = 2,color="red")+
      geom_text(aes(x = 10, y = 0.8, label = paste0(ros_2km,"% chance ROS > 2km/h")), fontface = "bold", size = 5)+
      labs(
        title = "Posterior Predictive Distribution of Rate of Spread",
        x = "Rate of Spread (km/h)",
        y = "Density",
        fontface="bold"
      ) +
      coord_cartesian(xlim = c(0, 20), ylim = c(0, 1)) +  # ← Adjust these as needed
      theme_minimal()+
  theme(
    text=element_text(face="bold",size=15)
  )
  })
}

shinyApp(ui, server)
```

