---
title: "Bayesian ROS model"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(here)

library(cmdstanr)
library(posterior)

library(ggplot2)
theme_set( theme_bw() )

MODEL_DIR <- here("stan_models")
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)

```

## Bayesian ROS model step by step

This doc loads ROS and weather data, and then builds a Bayesian ROS model step by step.


## Helper code

A utility function for sampling from cmdstanr models. This was provided in Richard McElreath's workshop materials (`script.R`) but has been slightly tweaked here to store the Stan model in specified directory and with a file name based on the name of the variable passed to the `model_code` argument. 

```{r}

get_samples <- function(model_code, data=list(), 
                        model_dir = MODEL_DIR, 
                        model_basename = NULL,
                        seed=123, chains=4) {
  
  if (is.null(model_basename)) {
    model_basename <- deparse(substitute(model_code))
    
    # remove '_code' from the base name if present
    model_basename <- sub(model_basename, pattern = "_code", replacement = "")
  }
  
  f <- write_stan_file(model_code, 
                       dir = model_dir, 
                       basename = model_basename)
  
  model_x <- cmdstan_model(f)
  samples_1 <- model_x$sample(
    data = data,
    seed = seed,
    chains = chains,
    parallel_chains = chains,
    refresh = 0
  )
  pr <- as_draws_rvars( samples_1$draws() )
  p <- list()
  for ( i in 1:length(pr) )
    p[[ names(pr)[i] ]] <- draws_of( pr[[i]] )
  return(p)
}

```




## ROS and weather

Load the ROS and weather data. Select a couple of weather variables to use as predictors for an initial model. Summarise the data

```{r}
dat.ros <- readRDS(here("data/lines.rds")) 
dat.weather <- readRDS(here("data/weatherhourly.rds"))

#select relevant ROS variable - ros and lineid, which matches lineid in weather data
dat.ros <- dat.ros %>% 
  select(lineid,ros=ros_kmh,
         start_time_utc,
         end_time_utc)

#select weather variables
#mean refers to mean of all barra cells that intersect the spread line.
dat.weather <- dat.weather %>% 
  select(lineid,
         wind=sfcWind_mean,
         rh=hurs_mean,
         time_utc_hourly
  ) %>% 
  
  mutate(wind=wind*3.6)


#weather is sampled hourly for the spread time of each line
#add variable to identify if BARRA sample is before, during or after spread time
dat.weather <- dat.weather %>% 
  left_join(dat.ros %>% select(lineid,start_time_utc,end_time_utc)) %>% 
  
  #round start and end times, to match barra times
  mutate(start_time_utc_round=lubridate::round_date(start_time_utc,unit = "hours"),
         end_time_utc_round=lubridate::round_date(end_time_utc,unit = "hours")) %>% 
  
  
  #remove any sampled barra weather from after end time
  filter(time_utc_hourly <= end_time_utc_round) %>% 
  
  #calculate time difference between barra sample time and start time
  #this may be useful to know, for example, wind speed 1 hour vs 3 hours before spread
  mutate(diff_start_hour=as.numeric(difftime(time_utc_hourly,start_time_utc_round,units = "hours"))) %>% 
  
  
  select(lineid,wind,rh,diff_start_hour)


#plot some of the weather and ros


dat.ros %>% 
  ggplot() +
  geom_histogram(aes(x=ros))

sample.ids <- sample(1:nrow(dat.ros), size = 9, replace = TRUE)
dat.weather %>% 
  filter(lineid %in% sample.ids) %>% 
  ggplot()+
  geom_histogram(aes(x=wind))+
  facet_wrap(~lineid)



```

### Example 1

Page 26 of workshop PDF.

You are told: "The windspeed is between 5 and 25 minutes"

A Stan model that represents this statement...

```{r}

model_code_1 <- "
parameters {
  real<lower=5, upper=6> windspeed;
  real<lower=1, upper=600>  temperature;
}
"

```


Even though this is a minimal model, we can compile it and derive posterior samples of the single parameter...

```{r}

# Write the model code to file
f <- write_stan_file(model_code_1, dir = MODEL_DIR, basename = "model_1")

# Compile the model (takes a minute)
model_1 <- cmdstan_model(f)

# Sample the model (this will be fast!)
samples_1 <- model_1$sample(
  data = list(),
  seed = 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 0
)

```


Inspect the samples.

```{r}

samples_1$summary()
post <- posterior::as_draws_df( samples_1$draws() )
plot(post$windspeed, xlab="sample", ylab="windspeed")
plot(density(post$temperature, adj=0.1), xlab="value", main="")

```



### Example 2

Page 34 of workshop PDF.

Lets say average wind speed is 9 kmh

A Stan model that represents this statement...

```{r}

model_code_2 <- "
parameters {
  real<lower=0, upper=20> windspeed;
}
model {
  windspeed ~ normal(10,2);
}
"

```


This time we will just use the helper function (defined at the top of this document) to compile and sample from the model.

```{r}

post2 <- get_samples(model_code_2)

```



```{r}

mu <- mean(post2$windspeed) |> round(2)
q90 <- quantile(post2$windspeed, c(0.05, 0.95)) |> round(2)

plot(density(post2$windspeed, from=0, adj = 0.1), 
     main = sprintf("Mean windspeed: %g\n90%% range: %g - %g", mu, q90[1], q90[2]), 
     xlab = "Windspeed")

```



### Example 3



You are given some data for observed wind speed and asked to estimate the average.

A Stan model for this task...


```{r}

model_code_3 <- "
data {
  int N;  // number of observations
  array[N] real<lower=0> y;  // observed values
}
parameters {
  // Unknown average value of windspeed
  real<lower=0.0001> avg_windspeed;
}
model {
  // Relate the observed values to the unknown (to be estimated)
  // average windspeed
  // Each value of y is assumed to come from this distribution
  y ~ exponential(1.0 / avg_windspeed);
}
"

```



```{r}
dat.test <- dat.weather %>% 
  filter(lineid <=10)

N=nrow(dat.test)
y=dat.test$wind


post3 <- get_samples(
  model_code_3,
  data=list(N=N, y=y)
)

```

```{r}

mu <- round(mean(post3$avg_windspeed), digits = 2)
q <- round(quantile(post3$avg_windspeed, c(0.05, 0.95)), digits = 2)

plot(density(post3$avg_windspeed, from=0, adj = 0.1), 
     main = sprintf("Average windspeed: unknown quanity - mean estimate: %g\n90%% range: %g - %g", 
                    mu, q[1], q[2]), 
     xlab = "Windspeed")

#add sample (data) mean
abline(v=mean(dat.test$wind))
```



### Example 4

You are given some data for windspeed observed at separate fire spread lines and asked to estimate the average for each location.

A Stan model for this task...

```{r}

model_code_4 <- "
data {
  int N;  // total number of observations
  int M;  // number of spread lines
  array[N] real<lower=0> windspeed;  // observed values
  array[N] real<lower=0> rh;  // observed values
  array[N] int x;  // location index (from 1) for each observation
}
parameters {
  // Unknown average value of waiting time at each location
  //Declare types, dimensions and constraints
  array[M] real<lower=0> avg_windspeed;
  array[M] real<lower=0> avg_rh;
}
model {
  // Relate the observed values to the unknown (to be estimated)
  // average duration. 
  // Note how we don't need an explicit loop here although we could do that 
  // for clarity at the expense of slightly longer computation time.
  // Also note that if we did 1.0/avg_windspeed[x] this would provoke an error
  // from Stan which is very picky about vector lengths etc.
  //
  windspeed ~ exponential(avg_windspeed[x]^(-1));
  rh ~ exponential(avg_rh[x]^(-1));
}
"

```


Define the data.

```{r}
dat.test <- dat.weather %>% 
  filter(lineid <=10) %>% 
  #give new lineid to ensure lineids start at 1
  mutate(newlineid=dense_rank(lineid))


N <- nrow(dat.test)

# spread line indices
x <- dat.test$newlineid

#Number of locations
M=n_distinct(dat.test$newlineid)

#wind speeds
y=dat.test$wind
rh=dat.test$rh


post4 <- get_samples(
  model_code_4,
  data = list(N=N, M=M, x=x, windspeed=y,rh=rh)
)


```




```{r}

# The `avg_....` elements in the `post4` object will be matrices with same N columns as unique lineids
# where each column has samples for a given line.

dat_gg <- as.data.frame(post4) %>% 
  select(-lp__) %>% 
  tidyr::pivot_longer(everything())

dat.sample.mean <- dat.test %>% 
  group_by(newlineid) %>% 
  summarise(avg_windspeed=mean(wind),avg_rh=mean(rh)) %>% 
  tidyr::pivot_longer(matches("avg")) %>% 
  mutate(name=paste0(name,".",newlineid))

ggplot(dat_gg, aes(x = value, colour = name)) +
  geom_density(linewidth = 1)+
  geom_vline(data=dat.sample.mean,aes(xintercept = value))+
  lims(x=c(0,50))+
  facet_wrap(~name)

```


### regression
a regression using summary weather values
```{r}
dat.test.2 <- dat.weather %>% 
  filter(diff_start_hour > -2) %>% 
  select(-diff_start_hour) %>% 
  group_by(lineid) %>% 
  summarise(mean_rh=mean(rh),mean_wind=mean(wind)) %>% 
  left_join(dat.ros)



```


```{r}

model_code_5 <- "
data {
  int<lower=0> N;   // number of data items/observations
  vector[N] wnd;           // first predictor - wind
  vector[N] rh;           // second predictor - relative humidity
  vector[N] ros;      // outcome - rate of spread
}
parameters {
  real alpha;           // intercept - could define weakly informative priors in model block
  real beta_wnd;       // coefficients for predictors
  real beta_rh;       // coefficients for predictors
  real<lower=0> sigma;  // error scale
}
model {
    ros ~ normal(alpha + beta_wnd * wnd + beta_rh * rh, sigma);  // likelihood. Assuming normal distrubtion for now, but this will need to change
}"


post5 <- get_samples(
  model_code_5,
  data = list(N=nrow(dat.test.2), 
              K=2,
              ros=dat.test.2$ros,
              wnd=dat.test.2$mean_wind,
              rh=dat.test.2$mean_rh)
)
```


```{r}
dat_gg <- as.data.frame(post5) %>% 
  select(-lp__) 


dat_gg %>% 
  tidyr::pivot_longer(everything()) %>% 
  ggplot(aes(x = value, colour = name)) +
  geom_density(linewidth = 1)+
  #lims(x=c(0,50))+
  facet_wrap(~name)
```

```{r}
#Create a grid of predictor values
wind_seq <- seq(0, 100, length.out = 100)
rh_mean <- mean(dat.test.2$mean_rh)


draws <- as.data.frame(post5)
# Get posterior mean estimates
beta_wnd <- draws$beta_wnd
beta_rh <- draws$beta_rh
alpha <- draws$alpha



#Generate posterior predictions for wind effect

# Use 100 draws for plotting, for speed
set.seed(123)
draw_idx <- sample(1:nrow(draws), 100)

pred_df <- expand.grid(
  wind = wind_seq,
  draw = draw_idx
) %>%
  mutate(
    alpha = alpha[draw],
    beta_wnd = beta_wnd[draw],
    beta_rh = beta_rh[draw],
    pred = alpha + beta_wnd * wind + beta_rh * rh_mean
  )





#Summarise the effect
library(tidyr)

pred_summary <- pred_df %>%
  group_by(wind) %>%
  summarise(
    mean = mean(pred),
    lower = quantile(pred, 0.05),
    upper = quantile(pred, 0.95)
  )

#plot
ggplot(pred_summary, aes(x = wind, y = mean)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  labs(
    x = "Wind Speed",
    y = "Predicted ROS",
    title = "Marginal Effect of Wind on ROS (RH fixed at mean)"
  )

```



